Temporal dataflows and their application to generative music

### 2017-03-23 article-icmc2017.tex
save work

    \subsection{Temporal formalism}
    Rappel i-score / ossia

@@

    Throughout this paper, we will use the execution traces in fig.~\ref{fig.simple} and fig.~\ref{fig.simple-reversed} as running example. 
    \begin{figure}
        \centering
        \begin{tikzpicture}[scale=0.5]
        \draw[line width=2pt] (0, 0) rectangle (5, 2);
        \draw[line width=2pt] (0 + 3, 0 - 2.5) rectangle (5 + 3, 2 - 2.5);

@@

        \node (t1) at (1.45, -3.45) {\large $t_1$};
        \node (t2) at (4, -3.45) {\large $t_2$};
        \node (t3) at (6.5, -3.45) {\large $t_3$};

@@

    \begin{figure}[h]
        \centering
        \begin{tikzpicture}[scale=0.5]

@@

    \section{Environment, implicit and explicit connections}
    In the traditional data flow paradigm, nodes are linked together through connections in their respective input and output ports.

    We extend this with a permanent external environment, that contains a mapping of key-value pairs. 
    Keys are specified with OSC-like\cite{osc} adresses: \lstinline|/foo/bar| and values can be any kind of data: numbers, strings, etc. 
    The values of the environment may vary independently of the execution of the program: for instance elements of the environment could be changed asynchronously through the network, graphical widgets or physical controls.

    We then have to provide a mean to use the values of this environment to the dataflow graph.
    Thus, we extend the inputs and outputs ports with the following expliciteness semantic:

    \begin{itemize}
        \item An explicit port has been connected by the developer to another port.
        \item An implicit port has not been connected to another port but may accept any number of adress - value pairs from the environment to perform its work.
    \end{itemize}

    During a tick, adresses have access to two sets of values: 
    \begin{itemize}
        \item The values that were in the environment at the beginning of the tick.
        \item The values produced by previous nodes in the graph.
    \end{itemize}

    It is up to the developer of nodes of the graph to choose whether a port can only work when connected explicitely, or with the environment produced from previous nodes, or with the global environment, or both.

    This mechanism allows for new patterns during the creation of a dataflow program: for instance, instead of accepting data from a single source, a port could specify a pattern (see OSC 1.1 paper with XPath) such as: \lstinline|/foo.*/bar/b[a-zA-Z]|. 
    % TODO pattern-matched addresses on the output: seems hard ? how to match it with a "pattern-matched" input port ??
	%Inner data: reduces the possibilities of connection: everything has to be there (explicit)

@@

    % TODO sample-precision of the activation / deactivation of a process / node
    % TODO 
	\section{Relationships}
    We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens.

     NOTE : if we have cables  we shouldn't have the "parameters" spill to other nodes. 
    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

@@

    We will get the following set of equations representing the behaviour: 
    \begin{itemize}
        \item During $t_0$: $\emptyset$.
        \item During $t_1$: $f_2 \circ f_1 $ or $f_1 \circ f_2$.
        \item During $t_2$: $\emptyset$. 
    \end{itemize}

@@

    The glutton execution may happen explicitely or implicitely.

@@

	The bufferisation can behave in two way:
	\begin{itemize}
		\item Readers of the buffer always start from the same point (the beginning of the previous function in the callback chain).
		\item Readers of the buffer continue from the latest read position.
		Question: what happens in case of multiple read ?
	\end{itemize}

    Note: this question has an answer in PLOP2006.

    The delayed execution has be specified explicitely for the tokens to be bufferised.

    If we have a delay connection from $f_1$ to $f_2$, the first call to $f_2$ will use the first value that was produced by $f_1$, and nothing before.

@@

    We propose for instance: 
    \begin{itemize}
        \item For simple parameters such as MIDI control changes or OSC messages: the latest value replaces the previous ones.
        \item For audio streams: the streams are mixed together.
        \item For MIDI notes: the notes are played together.
    \end{itemize}

@@

	\subsection{Replaced}
	All the nodes are run with the same input environment. 
	The resulting values put in the environment are those of the latest graph execution.

@@

    \section{Where to put the relationships}
    \subsection{In the cables}
    \subsection{In the ports}
    \subsection{In the nodes}
    \subsection{In the addresses}
    No temporal behaviour possible
    \subsection{Global behaviour}
    No temporal behaviour possible
    \subsection{Overall resolution}
    Unspecified by default.

@@

    \section{Default behaviours for orders}
    We study here the various behaviours possible to handle the dependency between nodes of the graph. 
    We can for instance leverage information provided in the temporal graph to connect the nodes at runtime.

    \subsection{Manual ordering}
    The user explicitely draws a cable between two ports.

    \subsection{Hierarchical}
    We follow the hierarchical organization of the temporal graph~: each time constraint has an ordered list of processes, and each time constraint is itself ordered in the scenario. This gives an order between processes.

    \subsection{Temporal}
    If no order is given, the temporal order in which objects are executed
    will become the order of chaining in the call graph.

    That is, if the execution happens in the order of the fig.~\ref{fig.simple}, we get the following execution:

@@

    Instead, if we are in the case of fig.~\ref{fig.simple-reversed}, we get for $t_1$ the execution $f_1 \circ f_2$.

	\section{Evaluation algorithm}
    We present here the evaluation algorithm that operates at each tick.

	Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration: see PLOP 2006 article (multi-stream, etc), sample accuracy, etc

>>>>

    \subsection{Temporal formalism}

    \begin{figure}
        \centering
        \includegraphics[width=0.45\textwidth]{images/iscore-example.eps}
        \caption{Part of an i-score scenario, showcasing the temporal syntax used. 
            A full horizontal line means that the time must not be interrupted, 
            while a dashed horizontal line means that the time of this time constraint can be interrupted to proceed 
            to the following parts of the score according to an external event.}
        \label{fig.iscore-example}
    \end{figure}

@@

    Throughout this paper, we will use the execution traces in fig.~\ref{fig.simple} and fig.~\ref{fig.simple-reversed} as running example.

    Finally, we will call developers the persons implementing dataflow-based software and users, authors, or composers the persons using dataflow-based software to implement their own music algorithms.

    \begin{figure}
        \centering
        \begin{tikzpicture}[scale=0.5]
        \draw[line width=2pt] (0, 0) rectangle (5, 2);
        \draw[line width=2pt] (0 + 3, 0 - 2.5) rectangle (5 + 3, 2 - 2.5);

@@

        \node (t1) at (1.45, -3.45) {\large $t_0$};
        \node (t2) at (4, -3.45) {\large $t_1$};
        \node (t3) at (6.5, -3.45) {\large $t_2$};

@@

    \begin{figure}
        \centering
        \begin{tikzpicture}[scale=0.5]

@@

    \section{Environment, implicit and explicit connections}
    In the data flow paradigm, nodes are linked together through connections in their respective input and output ports.

    We introduce a permanent external environment, that contains a mapping of key-value pairs.
    Keys are specified with OSC-like\cite{osc} adresses: \lstinline|/foo/bar| and values can be any kind of data: numbers, strings, etc. 
    The values of the environment may vary independently of the execution of the program: for instance elements of the environment may be changed asynchronously through the network, graphical widgets or physical controls.

    We then have to provide a mean to use the values of this environment to the dataflow.
    Thus, we extend the inputs and outputs ports with the following expliciteness semantic:

    \begin{itemize}
        \item An explicit port has been connected manually to another port.
        An input port may only read tokens from the output port it is connected to.

        \item An implicit port has not been connected to another port. 
        It may read and write to any number of adresses from the environment to perform its work.
        Every write done by a node is inserted in a sub-environment scoped to the duration of the tick.
    \end{itemize}

    During a tick, adresses have access to two sets of values: 
    \begin{itemize}
        \item The values that were in the environment at the beginning of the tick.
        \item The values produced by previous implicit output ports in the dataflow.
        Section~\ref{sec.relationships} presents the various policies that nodes can leverage to 
        use these values.
    \end{itemize}

    It is up to the implementor of nodes of the graph to choose whether a port can only work: 
    \begin{itemize}
        \item When connected explicitely,
        \item With the environment produced from previous nodes,
        \item With the global environment,
        \item With both of the previous.
    \end{itemize}

    This mechanism allows for new patterns during the creation of a dataflow program.
    For instance, instead of accepting data from a single source, a port could specify an input pattern (see OSC 1.1 paper with XPath) such as: \lstinline|/foo.*/bar/b[a-zA-Z]|.
    % TODO pattern-matched addresses on the output: seems hard ? how to match it with a "pattern-matched" input port ??
	%Inner data: reduces the possibilities of connection: everything has to be there (explicit)

@@

    % TODO sample-precision of the activation / deactivation of a process / node
    % TODO 
	\section{Relationships}
    \label{sec.relationships}
    We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens, whether the production is specified implicitely or explicitely.

    We propose five relationships : strict, glutton, delayed, mixed, and replaced. 
    For some of these relationships, an ordering on the nodes is necessary. 
    For this, a set of directed edges between nodes of the dataflow is used to provide this order.
    Methods to set up these edges are discussed in section~\ref{sec.order}.

    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

@@

    We will get the following set of equations representing the behaviour: 
    \begin{itemize}
        \item During $t_0$: $\emptyset$.
        \item During $t_1$: $f_2 \circ f_1 $ or $f_1 \circ f_2$ depending on the node ordering.
        \item During $t_2$: $\emptyset$. 
    \end{itemize}

@@

    The glutton execution may only happen when nodes are engaged in an implicit relationship.

@@

	The delayed execution can behave in two ways:
	\begin{itemize}
		\item Readers of the buffer always start from the same point : the beginning of the previous function in the callback chain.
		\item Readers of the buffer continue from the latest read position.
        In this case, the developer has to enforce that a single connection flowing from an output would be active at a given time, to prevent invalid reads .%Question: what happens in case of multiple read ?
	\end{itemize}

    %Note: this question has an answer in PLOP2006.

    The delayed execution has be specified explicitely for the tokens to be bufferised.
    It would be possible

    If we have a delay connection from $f_1$ to $f_2$, the first call to $f_2$ will use the first value that was produced by $f_1$.
    If the execution order happens to be reversed, such as in the case of fig.~\ref{fig.simple-reversed}, nothing will be read, and the execution will only happen during $t_1$.

@@

    We propose for instance the following operations: 
    \begin{itemize}
        \item For simple parameters such as MIDI control changes or OSC messages: the latest value replaces the previous ones.
        \item For audio streams: two alternatives : the streams are mixed together, or the streams are put in additional channels.
        \item For MIDI notes: the notes are played together.
    \end{itemize}

@@

	\subsection{Replaced}
	All the nodes are run with the same input environment. 
	The resulting values put in the environment are those of the latest node executing in the following dependent nodes.

    Software authors may still have to execute the previous nodes if they happen to produce side-effects.

@@

    \section{Specifying the relationships}
    A remaining problem is the specification of the kind of relationship between two ports, and, more precisely, 
    where this relationship should be specified.

    We study the implication of bearing such annotations for each object in the dataflow.

    \subsection{In the cables}
    The most precise way to set-up dependency-related information would be in 
    the cables connecting two ports.

    It is also the only way to set-up a delayed connection.

    However, it would also be the most tedious for users if the patch changes often.

    \subsection{In the ports}
    \subsection{In the nodes}
    \subsection{In the addresses}
    No temporal behaviour possible
    \subsection{Global behaviour}
    No temporal behaviour possible
    \subsection{Overall resolution}
    Unspecified by default.

@@

    \section{Default behaviours for orders}
    \label{sec.order}
    We study here the various behaviours possible to handle the dependency between nodes of the graph.
    information provided provided in the timeflow can be leveraged to connect the nodes at runtime in a fashion 
    that would make sense to composers \& authors.

    By default, any non-ordered node would strive to be scheduled at the earliest possible time.

    Connections between ports imply an ordering between nodes, hence we only consider the ordering between implicitely connected nodes of the dataflow.

    \subsection{Manual ordering}
    In this case, the author explicitely specifies an order by setting up an edge between two nodes. 
    This is the slowest process, but which gives the most precise specification.

    \subsection{Hierarchical}
    We follow the hierarchical organization of the timeflows~: each time constraint has an ordered list of processes, and each time constraint is itself ordered relative to others in a given timeflow. 
    This gives a natural order between processes.

    \subsection{Temporal}
    In this case, the temporal order in which objects are executed
    will become the order of chaining in the dataflow.

    That is, if a glutton execution happens in the order of the fig.~\ref{fig.simple}, we get the following execution:

@@

    Instead, if we are in the case of fig.~\ref{fig.simple-reversed}, we get:

    \begin{itemize}
        \item $t_0$: $f_2$.
        \item $t_1$: $f_1 \circ f_2$.
        \item $t_2$: $f_1$. 
    \end{itemize}

	%\section{Evaluation algorithm}
    %We present here the evaluation algorithm that operates at each tick.

	%Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration: see PLOP 2006 article (multi-stream, etc), sample accuracy, etc
